Experience Report: Haskell in Computational Biology

### 2012-05-08 icfp_reviewer_response.txt

We thank the reviewers for their insigful comments, and if our manuscript is 
accepted, we will incorporate changes that reflect these comments. We detail 
our responses to the reviewers below:

Reviewer A writes:

-- More discussion of the comparison between previous C++ software in 
-- that authors' research group and the new Haskell software would be 
-- much more interesting. In particular, the C++ tools described in 
-- section 4 seem significantly more complex than MRFy. Would Haskell 
-- scale to these problems just as well?

The C++ tools described are actually no more complex in terms of the algorithms 
they implement, though they comprise many more lines of code. We will clarify 
this language, and better contrast them.

-- Additionally, the performance discussion is so brief as to be almost 
-- uniterpretable. For example, it seems that MRFy parses protein 
-- structure (in what format?) in constant time, whereas SMURF seems to 
-- take longer on more complex proteins. Are the two tools implementing 
-- the same algorithm? Why is Haskell slower by a factor of 3 on the 
-- small structures? The discussion on parallelism is even briefer -- 
-- it's great to keep 48 cores warm, but what kind of parallel speedup is 
-- achieved?

We agree, and will provide more comprehensive quantitative results. In point of 
fact, MRFy is not implementing the same algorithm as SMURF, and we will better 
clarify this. We will also correct our wording to indicate that it's not 
*parsing* protein structures that we're talking about, but *computing* the 
optimal parse of a protein sequence onto a Markov random field. The complexity 
of the protein structures used to build the Markov random field is at issue 
here, and we will make this clearer. We will provide a better analysis of 
parallel performance, as well. In addition, while SMURF and MRFy differ 
algorithmically, we can and will provide a quantitative comparison of the 
run-time performance of the algorithm they share in common: Viterbi's algorithm.

-- The discussion of Viterbi's equations is extremely hard to follow for 
-- the uninitiated. This is especially true since the level of detail 
-- presented doesn't seem necessary for the larger points of the paper.

We agree, and given the request for more quantitative results, we will try to 
clarify the Viterbi discussion while at the same time going into less detail.

-- Is MRFy available? Can we try it out?

We plan to make MRFy publicly available by the time ICFP opens (and by which 
time we will have also published scientific results based on MRFy). We can make 
MRFy available for review if desired, but the primary investigator (Daniels' 
PhD advisor) is reluctant to release source code prior to publication of 
results.

-- I'm not sure that other structural biologists would agree that 
-- searching for homologous proteins is the "best known method of 
-- discovering" protein function.

We will clarify this to indicate that searching for homologous proteins is a 
well known and cost-effective approach that works in the absence of genetic 
interaction networks, given that protein sequence information is widely 
available. We will cite Pandey et al. 2007, "Computational Approaches for 
Protein Function Prediction: A Survey"

-- Why does the definition of HMMState on page 2 contain no data, just 
-- tags?

We realize that our use of "state" is misleading; our representation uses this 
HMMState data type as a tag, which along with node index and sequence residue 
index, allows us to index into our memoization table. We will change the data 
type to "StateLabel" and clarify this in the manuscript.

Reviewer B writes:

-- There is a report of positive experience using parallelism in Haskell, which is 
-- striking in its brevity. You could easily miss it! Apparently, at least for 
-- trivially parallel problems such as this, parallelism with GHC just works and 
-- there was nothing more to say. I'd like to see a bit more made of this result, 
-- since as a community we're often asked for evidence to support our hypothesis 
-- that functional programming is good for parallelism. Perhaps give parallelism a 
-- separate subsection and include a few numbers?

We plan to break parallelsm out into its own subsection, and include a 
quantitative analysis of scaling on several multi-core systems.

-- 3.1 
--    "data TProb" - why not use a newtype?

This is a historical artifact; TProb used to contain multiple fields. We will 
fix it.

-- 3.2 
--    - the code for "search" is a bit compressed - consider adding some line
--      breaks and blank lines to make it more readable

We will give "search" a bit more space.

-- 3.3 
--    - it sounds like your solution for vee' might be relying on lazy 
--      evaluation rather than properties of GHC's inliner, but it's hard 
--      to tell without more details.

We don't believe this to be the case, but it seems like a worthwhile and simple 
experiment. We will attempt to add strictness to vee' and report whether or not 
there is any impact.

-- 3.4 
--    - since version 7.4.1, GHC's profiler now annotates nested 
--      functions automatically.

This is terrific news! We plan to try GHC 7.4.1 once we resolve some library 
dependencies that are currently problematic. We will update the text to reflect 
this new support.

--   - it's not clear why the profiler could only be used on very small 
--     test cases: is it speed, memory use, or something else? If speed, 
--     does it help to use fewer annotations?

It was definitely runtime speed that was the problem here, for any test case 
large enough to be interesting, even if few annotations. We will state this 
more clearly and provide a more thorough evaluation of our profiling 
experiences.

-- 4. 
--    "working with HMMER codebase" -> "working with the HMMER codebase"

We will correct this.

Reviewer C writes:

-- The main thing I'm missing is more discussion of the effect of laziness -- in 
-- general and on the specific points mentioned. Was laziness useful for the 
-- algorithms described? Was it perhaps even instrumental? Or was it rather a 
-- hurdle (in particular with respect to the issues of debugging, profiling and 
-- performance)? Perhaps the authors can try to make some informed guess whether, 
-- with everything else being equal, a strict language might or might not have 
-- worked better for their problem domain.

We will include more discussion of laziness. We used laziness in two essential 
places, though one of them was not preserved: our original bottom-up 
dynamic-programming approach to Viterbi, and our use of a lazy infinite list of 
random numbers in our search implementation. We will discuss this in more 
detail.

In general, though, we don't particularly feel that laziness had a significant
bearing on our experience; other than one bug which we thought was due to
laziness, but actually turned out to be a mis-implemented typeclass derivation,
laziness was neither an impediment nor instrumental to our success.

-- Sec 3.1: In the type definition for Scored, what is type Score?

We will add this to section 3.1; as it happens it is "newtype Score = Score 
Double"

-- Does the code for vee' omit some cases, or what's happening for other 
-- HMMStates? Also, what is eProb?

Yes, this is only the Match case; we omitted the other recursive and base cases 
in the interests of brevity. We will enumerate the cases we've omitted, and 
refer the reader to Figure 3, but it would require too much space to include 
all of the code.

-- Sec 3.2: "Our implementation of hidden Markov models and Viterbi's algorithm 
-- duplicate existing functionality." I didn't get this. What functionality?

We will clarify this; we meant that we were reimplmenting some algorithms 
(namely, Viterbi) that have already been implemented (in C++) in SMURF and 
HMMER. The point of this contrast is to acknowledge that had we simply extended 
the SMURF code base to implement our new algorithm, we would not have had to 
reimplement Viterbi, for better or worse.

-- In the type definition for Scorer, what is type Placement?

In this case, [Int], but our real code is polymorphic. We ask the reviewers and 
committee: should we show the polymorphic version instead, though it may be 
less clear how it relates to this particular problem domain?

-- You say the use of higher-order functions was useful, e.g. for abstracting over 
-- search strategies. But isn't your type SearchStrategy effectively defining an 
-- object type? That is, wouldn't this have been as convenient with conventional 
-- object-oriented techniques?

We don't believe this would have been as convenient with conventional 
object-oriented techniques, as we are closing over a great deal of environment 
information (command-line parameters, some of which are search-strategy 
specific) which would have required instantiating various types of 
search-strategy-specific objects.

-- Sec 4: Typo: "Matt uses such data structures such as..." - double "such"

Indeed; we will fix this.

Reviewer D writes:

-- Perhaps the most useful feature of Haskell that is used here is function 
-- memoization which comes naturally to dynamic programming and is cumbersome in 
-- other programming languages save for the efficiency of the implementation of 
-- memoization in Haskell itself. It would have been useful to compare this to 
-- implementing table lookups in other programming languages.

We can provide a contrast to our experience implementing dynamic programming in 
C++ (unsurprisingly, there is a great deal more bookkeeping to be done). We do 
not feel we have the experience to provide a comparison over a wide range of 
programming languages.

-- While in the introduction the authors state that "considerations of performance 
-- are paramount" no numerical comparison is presented in the paper about 
-- performance of their implementations compared to others written in C or C++. 
-- Also, no viable way of comparing "convenience, readability, and productivity" 
-- to other implementations is presented in a quantitative manner that can help 
-- attract would-be Haskell programmers except for section on the use of 
-- memoization.

We promise a quantitative comparison should this manuscript be accepted. This 
comparison will largely have to be limited to code size and development time, 
but we can and will provide a quantitative comparison of the Viterbi algorithm 
run-time performance between SMURF and MRFy. However, the overall MRFy 
algorithm differs from SMURF, so overall run-time differences cannot be 
attributed to language choices as opposed to algorithmic differences.

-- Can the Authors identify other problems in computational biology, other than 
-- protein amino acid sequence comparison, where their experience presented in the 
-- paper would apply and be useful? I doubt it. For example in Section 5, "What 
-- can you learn from our experience?" The Authors state "If you are a 
-- computational biologist, you can be productive in Haskell, and you don’t need 
-- extensive experience." I would ask them to name one more problem in 
-- computational biology that would make use of their experience.

In fact, we can name several other problems we can (and may) address based on 
this experience. For example, we have applied graph-theoretic algorithms to the 
problems of gene functional annotation given a genetic interaction graph, as 
well as the problem of identifying compensatory genetic pathways 
("Between-pathway models") in the Yeast genome. We have written some Haskell 
code for the former problem (very much a work in progress), and while we 
approached the latter problem using Python, we feel that Haskell would have 
suited nicely. We will expand upon this discussion in the paper.

-- As stated earlier, identifying homologous proteins is a very important problem, 
-- but it is misleading to state that there are no other problems in computational 
-- biology that are important, the filed so very vast and problems so numerous 
-- which makes the title of the paper misleading. If the Conference Committee 
-- decides to accept the paper, I would urge them to ask the Authors to change the 
-- title of the paper.

If accepted, we will choose a more appropriately narrow title. It was not our 
intent to suggest that protein remote homology detection is the only field 
within computational biology, but rather that many computational biologists are 
unfamiliar with Haskell, and most computational biology software is written, as 
we suggest in the introduction, in C++, Java, or Python.

-- Also, In making a decision to accept or reject the paper, the Conference 
-- Committee should compare their decision with that of accepting or rejecting a 
-- paper that describes implementation of a dynamic programming algorithm in 
-- Hasekell. The phrase "computational biology" should not be a factor in their 
-- decision; the same methods presented in the paper could have been used in 
-- implementing an algorithm for comparing sequences of strings over a finite 
-- alphabet without change, regardless of the fact that in this case the alphabet 
-- encodes names of amino acids.

Clearly, we have not adequately explained the difference between our 
implementation (which combines dynamic programming, stochastic search, and a 
relatively new graphical model for Markov random fields, which are not the same 
as hidden Markov models) and past work from our group. We intend our experience 
report to be an exemplar of what computational biologists may experience should 
they choose to implement novel, performance-intensive algorithms in Haskell 
(or, perhaps, other functional languages). We will endeavor to make this goal 
clearer should our paper be accepted.

We would also like to better explain that MRFy represents a novel algorithm for 
solving Markov random fields, but whose components include implementations of 
the standard techniques of dynamic programming and stochastic search. We hope 
that some readers not yet familiar with Haskell will find this higher-order 
stochastic search implementation useful as well.

### 2012-05-08 icfp_reviewer_response.txt

Thanks for the help!  We're especially grateful for detailed questions
about such things as type definitions, which highlight issues we
overlooked.  We believe we can revise the paper to address most
concerns.  There are a couple of areas where reviewers may have
misunderstood our intent and we are not sure how best to revise the
paper.

As several reviewers recommend, we will present more comparison (and
more quantitative comparison) with previous C++ software.  We have few
quantitative measures that bear on "convenience, readability, and
productivity", but we will present the measures we have and will make
them easy to compare.  We will also discuss how Haskell might scale to
problems previously solved using C++.

We will provide a more thorough and more quantitative discussion of
performance, including measurements of parallel speedup and discussion
of parallelism.  Where possible we will present apples-to-apples
comparisons of Haskell and C++ codes.  Elsewhere we can compare
performance of Haskell code with the performance of C++ code on
different but related problems.

Reviewer A asks if MRFy is available.  The project PI (Daniels's PhD
supervisor) has agreed to release the source code before ICFP opens.
If reviewers want access now, we can provide a private copy.

Reviewer A asks why HMMState contains only tags and not data.  On
reflection, we think this type should not have been called a 'state'.
We plan to change the name, improve the presentation, and explain the
data type better.

Two reviewers ask about laziness.  We believe it was useful but not
instrumental.  To confirm, we will do and report on some experiments
with strictness annotations.

We will report in more detail that the profile was not useful because
it was too slow, even when using few annotations.

We will do our best to clear up the presentation of the Viterbi
algorithm and to characterize the cases we have omitted.

Reviewer C asks what functionality was duplicated in Section 3.2.
We will clarify: we meant that our Haskell code includes new
implementations of algorithms already implemented in C++ in both SMURF
and HMMER.  The revised paper will say more explicitly that this
reimplementation is a cost of using Haskell.  We suspect other
researchers may pay similar costs.

Reviewer C asks if SearchStrategy would have been as convenient using
conventional object-oriented techniques.  We will add the question to
the paper, with this answer: it is *easier* to create a SearchStrategy
in Haskell than using conventional object-oriented techniques, because
all the functions in a SearchStrategy are partial applications of
other functions.  In Haskell, capturing the partially applied
arguments is effortless; using conventional object-oriented
techniques, we would have to declare subclasses with instance
variables and would have to write coded to initialize objects of those
subclasses.

Reviewer D suggests comparing memoization using Data.Memocombinators
with table lookups in other languages.  We will contrast our Haskell
experience with our prior work implementing dynamic programming in C++.  
C++ requires much more bookkeeping.

Referee D asks for other problems in computational biology that would
make use of our experience.  We have applied Haskell to problems of
gene functional annotation given a genetic-interaction graph, using
standard graph algorithms.  We are also prototyping code to identify
compensatory genetic pathways ("Between-pathway models") in the yeast
genome.   Both of these efforts are much less mature than MRFy, and we
are not sure what, if anything, to say about them in the paper.

Referee D suggests the paper's title is too broad.  We certainly want
the title to reflect our experience accurately; if the paper does not
deliver on the promise implicit in the title, that's bad for us and
for ICFP.  But we would also like the paper to reach an audience
beyond the ICFP community.  In particular, if a graduate student in
computational biology wants to try Haskell, we would like this paper
to be usable as evidence.  We had hoped that a computational biologist
faced with a new problem would not take a short paper as
representative of all of computational biology, but instead would
evaluate the experience we describe and would decide whether the
evidence we present supports the use of Haskell for the new problem.

Perhaps a member of the program committee can suggest a title that is
accurate but is not so narrow as to rule out many problems in
computational biology?

### 2012-05-08 icfp_reviewer_response.txt

We will report in more detail that the profile was not useful because
it was too slow, even when using few annotations.

We will do our best to clear up the presentation of the Viterbi
algorithm and to characterize the cases we have omitted.

Reviewer C asks what functionality was duplicated in Section 3.2.
We will clarify: we meant that our Haskell code includes new
implementations of algorithms already implemented in C++ in both SMURF
and HMMER.  The revised paper will say more explicitly that this
reimplementation is a cost of using Haskell.  We suspect other
researchers may pay similar costs.

Reviewer C asks if SearchStrategy would have been as convenient using
conventional object-oriented techniques.  We will add the question to
the paper, with this answer: it is *easier* to create a SearchStrategy
in Haskell than using conventional object-oriented techniques, because
all the functions in a SearchStrategy are partial applications of
other functions.  In Haskell, capturing the partially applied
arguments is effortless; using conventional object-oriented
techniques, we would have to declare subclasses with instance
variables and would have to write code to initialize objects of those
subclasses.

@@

Referee D asks for other problems in computational biology that would
make use of our experience.  We have applied Haskell to problems of
gene functional annotation given a genetic-interaction graph, using
standard graph algorithms.  We are also prototyping code to identify
compensatory genetic pathways ("Between-pathway models") in the yeast
genome.   Both of these efforts are much less mature than MRFy, and we
are not sure what, if anything, to say about them in the paper.

Referee D suggests the paper's title is too broad.  We certainly want
the title to reflect our experience accurately; if the paper does not
deliver on the promise implicit in the title, that's bad for us and
for ICFP.  But we would also like the paper to reach an audience
beyond the ICFP community.  In particular, if a graduate student in
computational biology wants to try Haskell, we would like this paper
to be usable as evidence.  We had hoped that a computational biologist
faced with a new problem would not take a short paper as
representative of all of computational biology, but instead would
evaluate the experience we describe and would decide whether the
evidence we present supports the use of Haskell for the new problem.

Perhaps a member of the program committee can suggest a title that is
accurate but is not so narrow as to rule out many problems in
computational biology?

Reviewer D also implies that we have merely reimplemented a standard
dynamic programming algorithm.
Clearly, we have not adequately explained the difference between our 
implementation (which combines dynamic programming, stochastic search, and a 
relatively new graphical model for Markov random fields, which are not the same 
as hidden Markov models) and past work from our group. We intend our experience 
report to be an exemplar of what computational biologists may experience should 
they choose to implement novel, performance-intensive algorithms in Haskell 
(or, perhaps, other functional languages). We will endeavor to make this goal 
clearer should our paper be accepted.

We would also like to better explain that MRFy represents a novel algorithm for 
solving Markov random fields, but whose components include implementations of 
the standard techniques of dynamic programming and stochastic search. We hope 
that some readers not yet familiar with Haskell will find this higher-order 
stochastic search implementation useful as well.

### 2012-05-30 mrfy_experience_report.tex
fix caps
start of a note about quickcheck

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.Memocombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

@@

You~also need to know that while the Haskell community offers many
enticing tools, libraries, and packages,
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.Memocombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell the great packages from the wannabes
and the has-beens.
And even some of the great packages could be better documented, with
more examples.

>>>>

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

\remark{The following \P\ will need to be connected up to the last \S\
on ``our old approach to debugging.''}

Debugging led also not to awkward code, but to awkward interactions on
our team.
The two junior members of the team wanted to apply the debugging
skills they had honed through years of imperative programming.
The senior partner kept nattering on about QuickCheck properties.
No~progress was made until the first draft of this paper was prepared.
At~that point, we were quickly about to write QuickCheck
specifications for several properties of the Viterbi algorithm:
\begin{itemize}
\item
The total number of \verb+Mat+ and \verb+Ins+ states returned equals
the number of amino acids in the query sequence.
\item
The total number of matches, insertions, and deletions returned equals the
number of nodes in the hidden Markov model,
\emph{provided} that a maximal, nonempty, contiguous subsequence
of~\verb+Ins+ states counts as a single node,
and \verb+Mat+ and \verb+Del+ states count as one node each.
\item
The sequence returned satisfies the Plan7 invariant, that is, states
\verb+Ins+ and \verb+Del+ are never adjacent.
\end{itemize}
\textbf{These properties need to be followed by more ambitious
properties},
such as proving that nearby sequences are less likely than the
sequence returned by \verb+vee'+.

@@

You~also need to know that while the Haskell community offers many
enticing tools, libraries, and packages,
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.MemoCombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell the great packages from the wannabes
and the has-beens.
And even some of the great packages could be better documented, with
more examples.
